Making Sequential Programs Deterministic

### 2019-04-06 randoop-case-study.tex
Addd case study comments

To evaluate the usability of \theDeterminismChecker,
we applied it to the Randoop test
generator~\ifanonymous\citepalias{randoop-tool}\else\cite{PachecoLEB2007}\fi.
We also wrote specifications for libraries Randoop uses, such as the JDK, JUnit,
and others.
Like \theDeterminismCheckerImplementation,
all the case study materials are publicly available at \myurl{\todo{url for
    case study}} for reproducibility. \todo{This should include our link to
      our annotated plume-util, which we found was required to reproduce the
    results of the case study}

@@

We annotated one package at a time, starting with the most central packages
that are most depended upon.  Within a package, we followed a similar
strategy, annotating supertypes first.  If the determinism of classes and methods had been
documented, then our annotation effort
would have been easy, just converting English into type qualifiers.
Unfortunately, this was not the case: we had to reverse-engineer each
specification, largely from the methods it calls.  If a method $m$ calls a
method that we had previously annotated as nondeterministic, we annotated
$m$ as nondeterministic.
Sometimes after we
annotated a subclass, we had to go back and change annotations on
the supertype because the subclass implementation was
inconsistent with the specification we had guessed for the supertype.
After resolving most type-checking warnings by adding or changing
annotations, we manually examined each remaining one to determine whether
it was a bug or a false positive warning.
The effort would have been much easier for someone familiar with Randoop,
and yet easier if done while code is being written and is malleable.

@@

\begin{description}
\item[Severe issues]
  \strut
\begin{description}
\item[\bugHashSet]
One use of \<HashSet> could cause a problem if a type variable's lower or upper
bound has a type parameter that the type variable itself does not have.
This situation does occur, even in Randoop's test suite.
The developers fixed this by changing a \<HashSet> to \<LinkedHashSet>
(commit c975a9f7, shown in \cref{fig:randoop-bug-hashset}).
\TheDeterminismChecker confirmed that 
25 other uses of \<new HashSet> were acceptable, as were 15 uses of \<new HashMap>.

@@

\begin{itemize}
\item   (22\%)
%      8 constructors guarantee all instances of this class are @Det
  When every instance of a class is \<@Det>, \<@PolyDet> is equivalent to
  \<@Det>, but \theDeterminismChecker does not yet recognize this fact.
\item  (16\%)
%      6 https://github.com/t-rasmud/checker-framework/issues/96
  For a type that cannot be \<@OrderNonDet>, \<@PolyDet("up")> is
  equivalent to \<@PolyDet>.
\item (14\%)
%      5 can't specify receiver for anonymous class
  Java does not have syntax for specifying the type of an anonymous class.
\item (11\%)
%      4 actual type toString returns @PolyDet
  A variable is declared as \<Object> which has nondeterministic \<toString>,
  but at run time the value always has deterministic \<toString>.  For
  example, the run-time value is always an \<Integer> or a \<String>.
\item (5\%)
%      2 iterating over @OrderNonDet collection to modify another
  Iteration over an \<@OrderNonDet> collection in order to make another
  \<@OrderNonDet> collection, but the extracted elements have type
  \<@NonDet>.  For example, the following code is safe, but
  \theDeterminismChecker issues a warning.  Variable \<elt> has type
  \<@NonDet>, so the call to \<add> does not type check.
\begin{Verbatim}
  void m(@OrderNonDet List<@Det String> input) {
    @OrderNonDet List<@Det String> output = new List<>();
    for (String elt : input) {  // elt has type @NonDet String
      output.add(elt);          // error: illegal argument
    }
  }
\end{Verbatim}

@@

The first two issues can be resolved by improvements to
\theDeterminismCheckerImplementation.
The third and fourth issues are limitations of the Java language, but by a
local refactoring in the Randoop codebase can make the code type check.
Refactoring can also eliminate the last warning.

@@

In Randoop's 24K LOC of code, we wrote
98 \<@NonDet>,
64 \<@OrderNonDet>,
1083 \<@Det>,
and 103 \<@PolyDet> annotations.

>>>>

To evaluate the usability of \theDeterminismChecker,
we applied it to the Randoop test
generator~\ifanonymous\citepalias{randoop-tool}\else\cite{PachecoLEB2007}\fi.
We also wrote specifications for libraries Randoop uses, such as the JDK, JUnit,
and others.
Like \theDeterminismCheckerImplementation,
all the case study materials are publicly available at \myurl{\todo{url for
    case study}} for reproducibility. \todo{This should include our
      our annotated plume-util, which we found was required to reproduce the
    results of the case study}

@@

We annotated one package at a time, starting with the most central packages
that are most depended upon.  Within a package, we followed a similar
strategy, annotating supertypes first.  If the determinism of classes and methods had been
documented, then our annotation effort
would have been easy, just converting English into type qualifiers.
Unfortunately, this was not the case: we had to reverse-engineer each
specification, largely from the methods it calls.  If a method $m$ calls a
method that we had previously annotated as returning a nondeterministic result, we annotated
$m$ as returning a nondeterministic result.
Sometimes after we
annotated a subclass, we had to go back and change annotations on
the supertype because the subclass implementation was
inconsistent with the specification we had guessed for the supertype.
After resolving most type-checking warnings by adding or changing
annotations, we manually examined each remaining one to determine whether
it was a bug or a false positive warning.
\todo{In some cases, the number of \<@Det> annotations was especially prevalent.
  In these cases we changed the default qualifier for that class to \<@Det>.}
The effort would have been much easier for someone familiar with Randoop,
and yet easier if done while code is being written and is malleable.

@@

\begin{description}
\item[Severe issues]
  \strut
\begin{description}
\item[\bugHashSet]
One use of \<HashSet> could cause a problem if a type variable's lower or upper
bound in the code Randoop is run on has a type parameter that the type variable itself does not have.
This situation does occur, even in Randoop's test suite.
The developers fixed this by changing a \<HashSet> to \<LinkedHashSet>
(commit c975a9f7, shown in \cref{fig:randoop-bug-hashset}).
\TheDeterminismChecker confirmed that 
25 other uses of \<new HashSet> were acceptable, as were 15 uses of \<new HashMap>.

@@

\begin{itemize}
\item   (22\%)
%      8 constructors guarantee all instances of this class are @Det
  When every instance of a class is \<@Det>, \<@PolyDet> is equivalent to
  \<@Det>, but \theDeterminismChecker does not yet recognize this fact.
\item  (16\%)
%      6 https://github.com/t-rasmud/checker-framework/issues/96
  For a type that cannot be \<@OrderNonDet>, \<@PolyDet("up")> is
  equivalent to \<@PolyDet>.
\item (14\%)
%      5 can't specify receiver for anonymous class
  Java does not have syntax for specifying the receiver type in anonymous class
  methods.
\item (11\%)
%      4 actual type toString returns @PolyDet
  A variable is declared as \<Object> which has nondeterministic \<toString>,
  but at run time the subtype of \<Object> stored always has deterministic \<toString>.  For
  example, the run-time value is always an \<Integer> or a \<String>.
\item (5\%)
%      2 iterating over @OrderNonDet collection to modify another
  Iteration over an \<@OrderNonDet> collection in order to make another
  \<@OrderNonDet> collection, but the extracted elements have type
  \<@NonDet>.  For example, the following code is safe, but
  \theDeterminismChecker issues a warning.  Variable \<elt> has type
  \<@NonDet>, so the call to \<add> does not type check.
\begin{Verbatim}
  void m(@OrderNonDet List<@Det String> input) {
    @OrderNonDet List<@Det String> output = new List<>();
    for (String elt : input) {  // elt has type @NonDet String
      output.add(elt);          // error: illegal argument
    }
  }
\end{Verbatim}

@@

The first two issues can be resolved by improvements to
\theDeterminismCheckerImplementation.
The third and fourth issues are limitations of the Java language, but a
local refactoring in the Randoop codebase can make the code type check.
Refactoring can also eliminate the last warning. \todo{Can it? I thought
  what happens when iterating over an @OrderNonDet collection is something we
can't verify with the checker.}

@@

In Randoop's 24K lines of code, we wrote
98 \<@NonDet>,
64 \<@OrderNonDet>,
1083 \<@Det>,
and 103 \<@PolyDet> annotations.
